# Concluding Remarks

My writing was my humble attempt to convey my passion for the system of Biodanza and vibrating it to you. I hope that I was able to bring the message across  

- That life is deeply rooted in the universe

- That life is one

- That life is an emergent property that arises on different levels: from the chemical level, over cellular, tissue, organ, organ system, organism, community, ecosystem level upto the level of our entire planet, and that this also involves our human life that largely takes place in the symbolic social domain.

- That life is a gestalt of environment, cognition and an autopoietic unit (i.e. a cell, or organism, or ecosystem, ...), which can be viewed as an embodied mind.

- How each of these levels and thus also our psycho-socio-emotional realm share the same “bio-logic” or pattern of organisation, which invites us to deeply respect and to learn from all forms of life around us. 

- How we can broaden our vision on life and no longer reduce it to "biological life" only when, like Rolando, we define it as the two creative forces Anabasis and Catabasis.

These insights makes a biocentric world view very natural, which was Rolando's starting point when he developed his system of Biodanza. With Biodanza he envisioned to create an enriched environment with regular administration of stimuli through vivencia that can rebalance our biological system and can have a large impact on our ontogenesis. 

Indeed, vivencia has a strong impact on our body and brain. The sensations that we experience have a physiological impact through the modulation of our genes that encode for a diverse scala of proteins, hormones and neurotransmitters. 

Vivencia also have the quality that they immediately access our embodied mind through experience that precedes our extended consciousness. So it enables us to learn by strong experiences that directly tap into our personal, collective and vital unconsciousness. 

These transformative experiences through vivencia can be imprinted by epigenetics, which can induce important changes at a physiological level and also is a driver for mechanisms of learning and memory.  

Indeed, as Biodanza practitioners we are familiar with the long lasting memories to deep vivencia, and with how practicing Biodanza also induces change.
Change in how we feel, stand in life, interact socially and with our environment. 

I also hope that this monograph will trigger you to continue  on your Biodanza path with even more enthusiasm. Indeed, it is a unique path to promote your ontogenesis and well-being, and it invites you to reconnect with yourself, the others, and the entire world surrounding you. Moreover, it might also change the larger autopoietic social system in which you are embedded. And this, as Rolando Toro envisioned, by effectively switching it to another attractor through its positive feedback on your change, growth and interaction.

# Addendum {-} 

Below you will find some additional texts and impressions that I have written during my Biodanza teacher training.

## Reflection on my first Biodanza retreat with Annette Heynderickx in Buirefontaine {-}

Heel onverwachts en op de valreep verschenen we met zijn twee op Buirefontaine.   
En dat was alsof het zo moest zijn.   


Al meteen in de eerste sessie kregen we de kans om elkaar als koppel het mooiste cadeau te schenken:    
Onszelf,    
in zachtheid aangekomen,    
en in alle veiligheid opgenomen in een nieuwe groep,    
waardoor we elkaar volledig geopend en in volle afstemming    
opnieuw echt mochten ontmoeten.   


Een diepe intense ontmoeting,    
waarin we een hectisch jaar van zorg af mochten sluiten.   
Iets wat ons mentaal al langer duidelijk was,   
maar wat we op een of andere manier nog niet samen konden doorvoelen.   


En die ontmoeting liet dat in al zijn eenvoud en echtheid transformeren.   
Het was voor ons de kern van onze eerste vivencia (beleving),   
en tegelijk het begin van een intense reis die voor ons verscholen lag in het anagram die jij, Annette, legde.   


Tijdens volgende vivencia’s konden we in “de dans van de donder” de kracht ervaren in onszelf en die ook zien in de ander;   
om dan het mannelijke en vrouwelijke dat in ieder van ons schuilt te verkennen en te vieren;   
waarna de dans van de chaos ons uitnodigde om connectie te maken met het donkere in onszelf,   
dat donkere waar we zelf vaak blind voor zijn,   
en waar ook die scheppende kracht in verscholen ligt om als een kiem door te breken naar het nieuwe licht.   


En als je te midden van een vivencia even dreigde te overspoelen of te vervloeien,   
was er steeds die zorg, van jou, van jullie en van de volledige groep,   
zodat je die emotie, die beleving, die vivencia telkens opnieuw kon gaan containen.    
En dat gebeurde in alle eenvoud door een zorgende blik, een ontmoeting, een omhelsing, ...    

 
Dat is net de kracht van Biodanza,   
het nodigt je zo uit om te voelen, om af te stemmen en om te connecteren op een dieper weten waardoor je gaat transformeren.   
Het laat je versmelten met alle polen in jezelf,   
met je partner,   
met ieder uit de groep,   
en met "de elementen van het leven" die we dansten,   
waardoor je echt gaat ervaren,   
waardoor het denken voelen wordt,   
en daarna een innerlijk diep weten.   


En dat is een kunst die jij, Annette, als geen ander verstaat en weet over te brengen.    
Je lichaam spreekt,   
je dans nodigt uit om te beleven,   
jij en Frank scheppen de bedding van de groep met zoveel kracht,   
zodat ieder zichzelf en de ander mag en kan ervaren,   
zonder oordeel en zonder te veel woorden.   


De reis was heel inspirerend,   
het anagram werkt door ...    


Fien & Lieven - Juli 2020 



## Poetic impressions of some vivencia {-}

### De fysiologische loop {-}   

Mijn bewustzijn zakt   
van mijn hoofd   
in mijn lichaam.  
Ik voel en neem waar.  
Spanning wordt ontspanning.   
Ik laat los   
en beweeg.  
Het doen   
wordt langzaam  
zijn.  
  
```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat('
</br>'
)
```

### Ritmische coördinatie {-}   

Met mijn hand voel ik de ander  
en we worden samen één.  
Eén in de beweging,  
één met de beweging.   
Ik kleur jou   
en jij kleurt mij.  
We kleuren samen.  
En na de wissels   
regenbogen.  
  
```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat('  
</br>'
)
```

### Lopen met vastberadenheid {-}  

Ik voel de muziek   
en wordt geladen.  
Ik stuur mezelf naar voor,   
naar die overkant.   
En dan terug.   
  
Langzaam laat ik los.  
En bij de derde oversteek,  
wandel ik gewoon.  
Vastberaden en toch ontspannen.  
  
Ik voel die kracht en groei.  
Ik groei in grootsheid   
vanuit ontspanning.   
En dat Groot-Zijn   
zindert na.  
  
```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat('  
</br>'
)
```

### De ontmoeting {-}  

Vanuit dat Groot-Zijn    
zie ik jou.   
Ik zie jouw Groot-Zijn   
en jij die van mij.  
  
Stap voor stap   
kom je bij mij binnen   
en ik bij jou.   
  
Onze aura’s gaan versmelten   
en onze handen raken aan.  
Onze ogen vinden de weg  
heel diep bij elkaar naar binnen.   
Daar alwaar die puurheid   
in ons schuilt.  
  
Diep geraakt   
gaan we in omhelzing  
en we voelen na.   
  
```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat('  
</br>'
)
```

### Fluïditeit in compacte groep {-} 

Ik voel en beweeg.   
Ik ontmoet   
en ga verder.   
Ik raak aan  
en word aangeraakt.   
Steeds opnieuw  
en als maar verder.   
  
We bewegen en voelen   
we omhelzen en woelen  
samen zijn we één.  
  
We vloeien samen  
helemaal terug  
naar de oceaan.   
Intens aangekomen   
in onze oorsprong  
zijn we één.   
  
Muziek wordt zachter ...   
Van oceaan terug naar druppels.   
Vanuit het één worden we twee.   
En dan weer één.  
  
Je groeit uit mij   
in al je puurheid,  
daar zomaar in mijn oksel.   
En ik groei uit jou.  
Uit je armen en je handen,  
diep en zacht,   
en zo intens.  
  
Langzaam wiegend  
open ik mijn ogen en   
zie ik wie je bent.  
Langzaam wiegend   
open ik mijn ogen en   
zie ik wie ik ben.  
  
Daarna allen samen   
zo open en zo zacht   
In die intense cirkel   
schrijven we elk   
die nieuwe herinnering  
  
Een herinnering   
sterk en diep   
en o zo transformerend.  
De weg naar de oceaan   
is voor mij opnieuw geopend.   
  
De weg die wacht   
en roept   
van zodra muziek   
en dans weerklinkt.   
  
```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat('  
</br>' 
)
```

### Minotaurus {-}   

Jij Minotaurus,   
Toont ons het menszijn in al haar facetten.  
  
In het epicentrum van het veld van de aanwezige cirkel worden we herboren.  
Een zachte welp wordt een krachtige tijger.  
Een beer danst zijn angsten voor wilde dieren.   
Drenkelingen geven zich over en durven zich kopje onder te laten gaan om daarna het vertrouwen te vinden om zich te laten redden en het leven te omarmen.  
  
En ten midden van dit alles breng je voor mij de kracht van het loslaten,  
van niet te redden voor er wordt uitgereikt.  
De kracht van samen  
te zakken naar een eindeloze diepte  
En om daar dan te blijven staan ...   
  
Daar beneden  
word ik jij   
En jij wordt ik   
Omarmd dragen we elkaar   
terug naar de oppervlakte.  
Met jou Minotaurus   
als ons baken,  
een baken van liefde.   

<!--
## How quantum mechanics turned the Cartesian scientific world view upside down {-}

Quantum mechanics invites us to let go the old Cartesian way of causal thinking. 

In the Cartesian paradigm most processes are considered to be continuous and as soon as the initial conditions, such as position and velocity of particles, are known all future conditions can be derived from it. This led to our traditional deterministic world view and its associated cause and effect mind set. 

In quantum mechanics, however, energy can only be transferred in indivisible units, quanta, and we only know the probability that a quantum of energy will be transferred. Moreover, we do not now exactly when and where it will be transferred. The same holds for position and momentum^[Mass $\times$ Velocity] of matter. Particularly, probabilities for the possible results of measurements can be derived from the wave function that is used in quantum mechanics to represent matter. The theory further implies that both light and matter will display wave-like or particle-like characteristics depending on their interaction with their environment, which has been validated in many experiments. Moreover, also long range correlations have been shown to exist between so-called "entangled particles" that were separated by a large distance^[Quantum entanglement occurs when a group of "particles" are generated or interacted in a way such that the quantum state of each "particle" of the group cannot be described independently of the state of the others, even when the "particles" are subsequently separated by a large distance.]. So our world view can no longer remain deterministic, but, rather has to becomes probabilistic in nature. 

The quantum characteristics are especially observable on a microscopic level. On a macroscopic scale, however, processes typically involve so many quanta and/or an aggregate of so many "particles" that we can reason on the average energy and/or average position and momentum at this scale.  This, however, can be predicted in very narrow limits because we average over so many quanta or particles. Interestingly, the classical deterministic laws provide a good approximation for the average energy transfer and/or position and momentum of large matter aggregates. Moreover, a quantum is also a very tiny unit of energy so at a macroscopic scale the observed energy transfer involving many quanta also appears to be almost continuous. 

This clearly shows that scientific theories are ever-changing insights that are giving shape and form to how we view and experience the world; and more importantly that we have to remain open to change our mind set.

-->

